The Small World Network Effect in Software Project Teams \author{Kevin Peterson\\ \small \texttt{pete1968@umn.edu} }

### 2013-11-19 paper/main.tex
Added interview questions.

\title{
The Small World Network Effect in Open Source Project Teams
\author{Kevin Peterson\\
\small \texttt{pete1968@umn.edu}
}
}

@@

\noindent \\\textbf{Keywords.} Git, GitHub, Open Source

@@

A `small world' network is a social network characterized by high clustering of graph nodes paired with a short average path between nodes \cite{watts1998collective}. Milgram, in is seminal study, found that any two people are linked through a chain of friends and acquaintances on average of 6 people long\cite{milgram1967small}. From these observations, hypotheses can be made as to the efficiency of how knowledge is transfered in these groups\cite{latora2001efficient}. In this work, we investigate the `small world' phenomenon and how it relates to Open Source Project Teams.

\subsection{Hypothesis}

\textit{Open Source Project Teams will perform best at moderate levels of $Q$. This performance, measured by the number of interest their projects generate, will increase as the level of $Q$ increases for the developer graph. This will continue up to an optimal $Q$ value, after which performance will begin to degrade. Thus, the performance curve given $Q$ will be U-shaped, and match Uzzi's findings\cite{uzzi2005collaboration}. }

@@

\subsection{Collection and Storage}
Data was collected using the freely available FLOSSmole\cite{floss2006} data sets of free and open source projects. The data collection used was collected from the popular open source project site FreeCode\footnote{http://freecode.com/}, and contains data up to September, 2013.

\subsection{Analysis}
The collected data was organized into a graph structure, with the nodes representing the projects and the edges representing one or more shared developer between projects.

\subsubsection{Graph Structure}
A bipartite graph is a specialized type of graph containing two disjoint sets of verticies. This can be denoted by ${G=(P,C,E)}$, where ${P \cap C = \emptyset}$. For our purposes, let set $P$ be the set of all projects, and set $C$ be the set of all contributors. Each project node, therefore, is only connected to contributor nodes, and vice versa, as show in figure \ref{fig:example_bipartite_graph}. This bipartite representation is typical of what is found in social and collaboration networks\cite{ramasco2004self}.

@@

@@

\subsubsection{Graph Analysis}
To support our hypotheses, there are several key metrics to be studied in regards to the aforementioned graph. Many of these metrics follow the work of Boccaletti \textit{et. al}\cite{boccaletti2006complex}, and other studies of similar graph types CITATION HERE.

\textit{Clustering Coefficient (Transitivity)}\\
We calculate the \textit{Clustering Coefficient} (or $C^\Delta$) as a measure of the proportion of closed triangles in a graph\cite{newman2003structure}:
\[C^\delta = 3 \times \frac{\text{number of triangles}}
                    {\text{number of connected triples}}\]

\textit{Clustering Coefficient (Average)}\\
Another approach to clustering is to take the average of each node's local clustering coefficient\cite{watts1998collective}.

\[ c_i = \frac{|\{e_{jk}\}|}{k_i(k_i-1)} :\, v_j,v_k \in N_i,\, e_{jk} \in E \]

@@

\textit{Average Path Length}\\
Also, calculating the \textit{Path Length} ($L$) allows us to determine how far removed developers in the graph are from one another. The path length is calculated as the average distance between any two node pairs in the graph:

@@

Results were analyzed using the Python graph processing packaged NetworkX\cite{hagberg-2008-exploring}.

@@

  \draw ({n1}) -- ({n2});
  \draw ({n3}) -- ({n4});
  \draw ({n1}) -- ({n4});
  \draw ({n2}) -- ({n4});
\end{tikzpicture}
\caption{A caption.}
\label{fig:example_graphs}
\end{figure}

\textit{Small-World-ness}\\
The idea of a \textit{small world} can be formalized by the following equations\cite{humphries2008network}.

@@

\textit{Successful vs Unsuccessful}\\
To further explore the data, a comparion was made between \textit{successful} and \textit{unsuccessful} projects was made. We define \textit{successful} as any project with {$PopularityScore > 100$}, and \textit{unsuccessful} as any project with {$PopularityScore < 50$}. Given this view of the data, we can then make assertions about how we expect these types of projects to behave, given the stated hypothesis. More specifically, we can examine not only if the u-shaped curve holds for \textit{popular} projects, but if the inverse holds for \textit{unsuccessful} ones.

\subsubsection{Qualitative Analysis}
It has been asserted that people tend to make connections with people with whom they share similarities\cite{mcpherson2001birds}. This idea of homophily, however, needs to be further expanded before we can make assertions about correlations to performance. It has been shown that demographic homophily is complex to measure, correlates little to performance outcomes, and in fact correlate negatively\cite{reagans2004make,lawrence1997perspective}. For our purposes, it is much more appropriate to examine how teams best leverage similarities of knowledge and information, as the flow of these assets can be a key driver to success\cite{nissen2002extended}.

@@

\begin{figure}
\begin{center}
\caption{$C^\Delta$ and Project Popularity}
\includegraphics[width=0.5\textwidth]{images/freecode-graph.png}
\end{center}
\caption{A caption.}
\label{fig:cc_graph}
\end{figure}

@@

\begin{figure}
\begin{center}
\caption{$C^\Delta$ of Popular Projects}
\includegraphics[width=0.5\textwidth]{images/freecode-popular.png}
\end{center}
\caption{A caption.}
\label{fig:cc_popular}
\end{figure}

@@

\begin{figure}
\begin{center}
\caption{$C^\Delta$ of Unpopular Projects}
\includegraphics[width=0.5\textwidth]{images/freecode-unpopular.png}
\end{center}
\caption{A caption.}
\label{fig:cc_unpopular}
\end{figure}

\begin{figure}
\begin{center}
\caption{$Q$ Distribution}
\includegraphics[width=0.5\textwidth]{images/freecode-q-histo.png}
\end{center}
\caption{A caption.}
\label{fig:q_distribution}
\end{figure}

\begin{figure}
\begin{center}
\caption{$S^\Delta$ Distribution}
\includegraphics[width=0.5\textwidth]{images/freecode-smallworld-histo.png}
\end{center}
\caption{A caption.}
\label{fig:sw_distribution}
\end{figure}

The project graph contains a large amount of subgraphs, and as a whole appears less clustered than some other observed social network graphs\cite{madey2002open}.

@@

\section{Conclusion}

>>>>

\title{
The Small World Network Effect in Software Project Teams
\author{Kevin Peterson\\
\small \texttt{pete1968@umn.edu}
}
}

@@

\noindent \\\textbf{Keywords.} Small World, Project Management, Software

@@

A `small world' network is a social network characterized by high clustering of graph nodes paired with a short average path between nodes \cite{watts1998collective}. `Clustering' in this context means how closely nodes in the graph are related to each other, or "the friend of my friend is also my friend." Milgram, in is seminal study, found that any two people are linked through a chain of friends and acquaintances on average of 6 people long\cite{milgram1967small}. From these observations, hypotheses can be made as to the efficiency of how knowledge is transfered in these groups\cite{latora2001efficient}. In this work, we investigate the `small world' phenomenon and how it relates to Software Project Teams.

\subsection{Hypothesis}
The formalization of the `small world' quality of a graph can be expressed by the \textit{Small World Quotient}, or $Q$\cite{watts1999small,watts1998collective}. The value $Q$ is calculated by dividing the clustering factor of graph by the average path length between nodes. Uzzi explored this value in graphs of Broadway productions\cite{uzzi2005collaboration}. His findings suggest that the success of Broadway productions is greatest when teams have a good mix of new and familiar members. Productions are less likely to be successful when team members are very familiar with each other. Also, probability of success drops much the same way when there is little familiarity between team members. This indicates that there is an optimal ratio of new and familiar members on a team -- or, an optimal value of $Q$ when related to performance.

\textit{Software Project Teams will perform best at moderate levels of $Q$. This performance, measured by the number of interest their projects generate, will increase as the level of $Q$ increases for the developer graph. This will continue up to an optimal $Q$ value, after which performance will begin to degrade. Thus, the performance curve given $Q$ will be U-shaped, and match Uzzi's findings\cite{uzzi2005collaboration}.}

@@

\subsection{Collection and Storage}
Data was collected using the freely available FLOSSmole\cite{floss2006} data sets of free and open source projects. The data collection used was collected from the popular open source project site FreeCode\footnote{http://freecode.com/}, and contains data up to September, 2013. Results were analyzed using the Python graph processing packaged NetworkX\cite{hagberg-2008-exploring}. All source code pertaining to the collection and analysis can be found at \url{https://github.com/kevinpeterson/small-world-effect-research}.

\subsection{Analysis}
The collected data was organized into a graph structure, with the nodes representing the projects and the edges representing one or more shared developer between projects. Various properties of the graph are analyzied, as indicated in the sections below.

\subsubsection{Graph Structure}
We first consider the FreeCode FLOSSmole\cite{floss2006} data sets to be one large graph. From this large disconnected graph, each connected subgraph is analyzed. Metrics are computed for each of these subgraphs, and a total subgraph performance is computed by averaging the performance value of each graph in the subgraph. We measure a subgraph's perfomance by the FreeCode \textit{Popularity Score} metric, which is calculated the following formula\footnote{http://help.freecode.com/kb/statistics/how-do-you-measure-a-projects-popularity}:

\[ ((record\ hits + URL\ hits) \cdot (followers + 1))^{1/2} \]

@@

A bipartite graph is a specialized type of graph containing two disjoint sets of verticies. This can be denoted by ${G=(P,C,E)}$, where ${P \cap C = \emptyset}$. For our purposes, let set $P$ be the set of all projects, and set $C$ be the set of all contributors. Each project node, therefore, is only connected to contributor nodes, and vice versa, as show in figure \ref{fig:example_bipartite_graph}. This bipartite representation is typical of what is found in social and collaboration networks\cite{ramasco2004self}.

@@

\subsubsection{Graph Analysis}
To support our hypotheses, there are several key metrics to be studied in regards to the aforementioned graph. Many of these metrics follow the work of Boccaletti \textit{et. al}\cite{boccaletti2006complex}, and other studies of similar graph types\cite{latora2001efficient,adamic1999small}.

\noindent\\\textit{Clustering Coefficient - Transitivity ($C^\Delta$)}\\
We calculate the \textit{Clustering Coefficient} (or $C^\Delta$) as a measure of the proportion of closed triangles in a graph\cite{newman2003structure}:
\[C^\Delta = 3 \times \frac{\text{number of triangles}}
                    {\text{number of connected triples}}\]

\noindent\\\textit{Clustering Coefficient - Average ($\overline{C^\Delta}$)}\\
Another approach to clustering is to take the average of each node's local clustering coefficient\cite{watts1998collective}.

\[ C_i = \frac{|\{e_{jk}\}|}{k_i(k_i-1)} :\, v_j,v_k \in N_i,\, e_{jk} \in E \]

@@

\[ \overline{C^\Delta} = \frac{1}{|P|}\sum^{|P|}_{i=1}C_i \]
Unless otherwise noted, the Transitivity version of this metric ($C^\Delta$) will be used for all further calculations and metrics.

\noindent\\\textit{Average Path Length ($L$)}\\
Also, calculating the \textit{Path Length} ($L$) allows us to determine how far removed developers in the graph are from one another. The path length is calculated as the average distance between any two node pairs in the graph:

@@

\noindent\\\textit{Small World Quotient ($Q$)}\\
The \textit{Small World Quotient} ($Q$) is defined as the Clustering Coefficient ($C^\Delta$) divided by the  Average Path Length ($L$)\cite{watts1999small,watts1998collective}.

\[Q =  \frac{C^\Delta}{L}\]

@@

  \draw ({n1}) -- ({n2});
  \draw ({n3}) -- ({n4});
  \draw ({n1}) -- ({n4});
  \draw ({n2}) -- ({n4});
\end{tikzpicture}
\caption{Example Clustering Coefficient ($C^\Delta$) values for sample graphs}
\label{fig:example_graphs}
\end{figure}

\noindent\\\textit{Small-World-ness ($S^\Delta$)}\\
The definition of a \textit{small world} graph can be formalized by the following equations\cite{humphries2008network}.

@@

\noindent\\\textit{Successful vs Unsuccessful}\\
To further explore the data, a comparion was made between \textit{successful} and \textit{unsuccessful} projects was made. We define \textit{successful} as any project with {$PopularityScore > 100$}, and \textit{unsuccessful} as any project with {$PopularityScore < 50$}. Given this view of the data, we can then make assertions about how we expect these types of projects to behave, given the stated hypothesis. More specifically, we can examine not only if the u-shaped curve holds for \textit{popular} projects, but if the inverse holds for \textit{unsuccessful} ones.

\subsubsection{Qualitative Analysis}
It has been asserted that people tend to make connections with people with whom they share similarities\cite{mcpherson2001birds}. This idea of homophily, however, needs to be further expanded before we can make assertions about correlations to performance. It has been shown that demographic homophily is complex to measure, correlates little to performance outcomes, and in fact correlate negatively\cite{reagans2004make,lawrence1997perspective}. For our purposes, it is much more appropriate to examine how teams best leverage similarities of knowledge and information, as the flow of these assets can be a key driver to success\cite{nissen2002extended}.\\
In the same way we can explore the role of team familiarity. The two types of familiarity that we will explore are \textit{member} familiarity and \textit{task/project} familiarity\cite{harrison2003time}. These metrics, although not captured in data, could be important factors in \textit{how} teams form, and thus help us to understand \textit{why} teams perform the way that they do.\\

To further study these factors, we arranged interviews with two current professionals in the field. Both have had experience with project management and are certified Project Management Professionals (PMP)\textregistered. We proposed to them the following questions:

\noindent\\\textit{Q1: How do development teams collaborate and share ideas?}\\
\noindent\\\textit{Q2: Do teams generally remain intact over multiple projects, or do members often shuffle and re-combine?}\\
\noindent\\\textit{Q3: When assembling a team, which is preferred -- members familiar with each other, or members familiar with the task or project.}\\
\noindent\\\textit{Q4: What are some characteristics of a high performing team?}\\
\noindent\\\textit{Q5: In your experience, is it better to (A) keep an innovative team intact, (B) periodically bring in new members or re-assign exising members, or (C) split up and disperse the team to as many other projects as possible?}\\

@@

\begin{figure}
\begin{center}
\includegraphics[width=0.5\textwidth]{images/freecode-graph.png}
\end{center}
\caption{$C^\Delta$ and Project Popularity}
\label{fig:cc_graph}
\end{figure}

@@

\begin{figure}
\begin{center}
\includegraphics[width=0.5\textwidth]{images/freecode-popular.png}
\end{center}
\caption{$C^\Delta$ of Popular Projects}
\label{fig:cc_popular}
\end{figure}

@@

\begin{figure}
\begin{center}
\includegraphics[width=0.5\textwidth]{images/freecode-unpopular.png}
\end{center}
\caption{$C^\Delta$ of Unpopular Projects}
\label{fig:cc_unpopular}
\end{figure}

Distribution of $Q$ values are shown in figure\ref{fig:q_distribution}. This indicates that the majority of sub-graphs have a $Q$ value near either 0.0 or 1.0.

\begin{figure}
\begin{center}
\includegraphics[width=0.5\textwidth]{images/freecode-q-histo.png}
\end{center}
\caption{$Q$ Distribution}
\label{fig:q_distribution}
\end{figure}

Small-world-ness, or $S^\Delta$, is generally between 1 and 3 for most sub-graphs.

\begin{figure}
\begin{center}
\includegraphics[width=0.5\textwidth]{images/freecode-smallworld-histo.png}
\end{center}
\caption{$S^\Delta$ Distribution}
\label{fig:sw_distribution}
\end{figure}

@@

\section{Conclusion}
The project graph contains a large amount of subgraphs, and as a whole appears less clustered than some other observed social network graphs\cite{madey2002open}.
